---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tblcleanr

<!-- badges: start -->
<!-- badges: end -->

The goal of tblcleanr is to make it easier to tidy real world data tables.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("trekonom/tblcleanr")
```

## Example

Here is a typical example of how to use `tblcleanr` to tidy a messy data table. 
The example data is drawn from the [`unpivotr`](https://nacnudus.github.io/unpivotr/) 
package which serves a similar purpose and follows a simliar approach as `tblcleanr`:

```{r}
library(tidyverse)
library(unpivotr)
library(tblcleanr)

x <- purpose$`up-left left-up`

head(x)

y <- x |>
  # Fill empty cells in first header row
  fill_row(1) |>
  # Fill qualification column
  fill(1) |>
  # Move header rows to colum names
  rows_to_names(1:2) |>
  # Rename tidy columns
  rename(qualification = 1, "age-band" = 2) |>
  # Reshape to tidy format
  pivot_longer(-c(qualification, `age-band`),
    names_to = c("sex", "life_satisfaction"),
    names_sep = "_",
    values_to = "count"
  ) |> 
  # Convert data type
  mutate(count = as.integer(count))

head(y)
```
