---
title: "Get started with tblcleanr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with tblcleanr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tblcleanr)
```

Real-world spreadsheets are rarely tidy: header information is spread
across several rows, category labels sit in their own rows instead of
their own column, and the row/column names read.csv()/readxl generate
(`X2`, `X3`, ...) leak into the data. `tblcleanr` provides small,
composable functions to turn tables like that into tidy data, one row
operation at a time.

This vignette walks through a typical example: a cross-tabulation with
a two-row header (sex, then age band) and a leading label column, read
in the shape a spreadsheet import would produce it.

```{r}
x <- data.frame(
  X2 = c(NA, NA, "Bachelor's degree", NA, NA, NA),
  X3 = c(NA, NA, "15 - 24", "25 - 44", "45 - 64", "65+"),
  X4 = c("Female", "0 - 6", "7000", "12000", "10000", NA),
  X5 = c(NA, "7 - 10", "27000", "137000", "64000", "18000"),
  X6 = c("Male", "0 - 6", NA, "9000", "7000", "7000"),
  X7 = c(NA, "7 - 10", "13000", "81000", "66000", "17000")
)
x
```

Row 1 holds the `sex` header, but only where it changes ("Female" only
appears once, "Male" only once); row 2 holds the `life_satisfaction`
band; row 3 onwards holds the `qualification` label in the first
column (only on its first occurrence) and the actual counts.

## Fill in the gaps in the header rows

[`fill_row()`](../reference/fill_row.html) fills `NA`s across a row,
the row-wise equivalent of `tidyr::fill()`. Use it to propagate
`"Female"`/`"Male"` across the columns they apply to:

```{r}
x <- fill_row(x, 1)
x
```

## Fill in the gaps in a column

The `qualification` label in `X2` only appears once per group, in the
first data row of each group — that's `tidyr::fill()`, applied to a
column rather than a row:

```{r}
x <- tidyr::fill(x, X2)
x
```

## Promote header rows to column names

[`rows_to_names()`](../reference/rows_to_names.html) turns one or more
rows into column names, concatenating them with `sep`. By default it
also renames any column whose name looks like a spreadsheet-import
artifact (`X2`, `X3`, ... or `...1`, `...2`, ...), which is why `X2`
picked up a name here too even though rows 1:2 are blank in that
column:

```{r}
x <- rows_to_names(x, 1:2)
x
```

## Rename and reshape to tidy form

From here on, `x` is a normal (if wide) data frame, so ordinary
dplyr/tidyr verbs take over: rename the still-generic columns, then
`pivot_longer()` the sex/life-satisfaction combinations into rows.

```{r}
x <- dplyr::rename(x, qualification = 1, "age-band" = 2)
x <- tidyr::pivot_longer(
  x,
  -c(qualification, `age-band`),
  names_to = c("sex", "life_satisfaction"),
  names_sep = "_",
  values_to = "count"
)
x <- dplyr::mutate(x, count = as.integer(count))
x
```

## Other building blocks

`tblcleanr` has a few more row operations for shapes that show up
often in messy spreadsheets, each documented with a runnable example:

- [`collapse_row()`](../reference/collapse_row.html) concatenates
  several rows into one, e.g. when a header is split across two rows
  that should become a single label rather than two separate ones.
- [`category_row_na()`](../reference/category-row-to-column.html) and
  [`category_row_regex()`](../reference/category-row-to-column.html)
  move a category label that sits in its own row (either detected by
  "the rest of the row is empty" or by a regex) into a proper column.
- [`remove_rows_not_na_n()`](../reference/remove_rows_not_na_n.html)
  and friends drop rows/columns based on how many missing values they
  have, e.g. to remove separator or footnote rows.
- [`str_replace_row()`](../reference/str_replace_row.html) and
  [`str_remove_row()`](../reference/str_remove_row.html) run a
  `gsub()`-style replacement across specific rows only, e.g. to strip
  footnote markers.
- [`as_excel_date()`](../reference/as_excel_date.html) converts Excel's
  numeric date serials to proper R `Date`s.
